\PIII, linearizing unit cell parameters

### 2025-11-23 P3_uv5038/P3_uv5038.tex
just saving work

	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
	\author[b]{Herbert J.}{Bernstein}

\aff[a]{Ronin Institute for Independent Scholarship 2.0, USA}
\aff[b]{Ronin Institute for Independent Scholarship 2.0, USA}

@@

	\begin{abstract}
		The space \PIII{} is introduced, derived from unit cell axial lengths and interaxial angles. \PIII enables linearization of unit cell parameters using three polar coordinate bases.

@@

	For the above reasons, other spaces are often used to describe lattices, often for specialized purposes; see Table  \ref{tbl_spaces}. Figure \ref{fig_spaces} shows relationships between
	several of these spaces. Although some of these spaces, such as \HVI{} and	\FIII{} are in almost daily use, they are not often referred to 
	as ``spaces''. In the case of \FIII{}, symmetry elements are linear
	operations in this space, and fractional coordinates of atomic positions
	are vectors expressed in this space.

@@

	} &
	Composite geometric parameter space $\mathbb{R}_{>0}^3 \times (0^\circ, 180^\circ)^3$. Lengths are unconstrained; angles are geometrically restricted. The sum of the angles must be less than $360^\circ$ and obey the triangle inequality \HVI{} is useful for conventional lattice descriptions, but limited in symmetry representation and metric-based classification. \\
	\midrule
	\FIII{} &
	$\vec{a},\, \vec{b},\, \vec{c} \in \mathbb{R}^3$ &
	Base vector space in \RIII. Coordinates are the three-space base vectors of a unit cell. Lattices are defined using the
	components of \FIII{}, and translational symmetry is described 
	in \FIII{}. \\	
	\midrule

@@

	\PIII{} &
	\parbox[t]{5.5cm}{
		$(|\vec{a}|,\alpha),\ (|\vec{b}|,\beta),\ (|\vec{c}|,\gamma)$\\
		$\Rightarrow$\\
		\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha)$\\
		\hspace*{0.5cm}$(|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta)$\\
		\hspace*{0.5cm}$(|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$\\
		also known as $(p_1, p_2, p_3)$
	} &
	Polar coordinate space defined in terms to unit cell axial
	lengths and their opposing angle. It is a compact and smooth alternative to $\HVI$. Useful for comparing unit cells. (This work) \\
		\midrule

	\SVI{} &
	\parbox[t]{5.5cm}{
		$s_1 = \vec{b} \cdot \vec{c},\, s_2 = \vec{a} \cdot \vec{c},\, s_3 = \vec{a} \cdot \vec{b}$\\
		$s_4 = \vec{a} \cdot \vec{d},\, s_5 = \vec{b} \cdot \vec{d},\, s_6 = \vec{c} \cdot \vec{d}$\\
		\quad where $\vec{d} = -\vec{a} - \vec{b} - \vec{c}$
	} &
	Used for Selling/Delone reduction. The scalar components are dot products between lattice vectors. \SVI{} supports Bravsi lattice classification. 24 Bravais types were defined using the scalar components of \SVI{} by Delone (1975), later revised using \SVI{} \cite{andrews2023measuring}. \\
		\midrule

@@

\CIII{} &
	\parbox[t]{5.5cm}{
		From \SVI{} scalars: $s_1,\dots,s_6$\\
		$c_1 = (s_1, s_4),\ c_2 = (s_2, s_5),\ c_3 = (s_3, s_6)$\\
		where the parentheses denote a complex number
	} &
	\CIII{} associates scalar pairs identified by Delone as “opposites.” Symmetry and Selling/Delone reduction simplify in this space \cite{andrews2023complex}. \\
		\midrule

@@

\begin{figure}
	\centering
\begin{tikzpicture}[
	node distance=1.2cm and 2.0cm,
	coordspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
		fill=gray!20, rounded corners, font=\small, thick},
	metricspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
		fill=blue!10, rounded corners, font=\small, thick},
	derivedspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
		fill=green!10, rounded corners, font=\small, thick},
	arrowlabel/.style={font=\scriptsize, midway, fill=white, inner sep=1pt}
	]

	% Row 1: Fundamental space
	\node[coordspace] (F3) {\FIII{} \\ Base vectors};

	% Row 2: Scalar & parameter spaces
	\node[metricspace] (G6) [below left=of F3] {\GVI{} \\ Metric tensor};
	\node[metricspace] (S6) [below=of F3] {\SVI{} \\ Selling/Delone};
	\node[coordspace] (H6) [below right=of F3] {\HVI \\ Cell params};

	% Row 3: Derived/reduced spaces
	\node[derivedspace] (DC7u) [below=of G6] {DC7u \\ Dirichlet coords};
	\node[derivedspace] (C3) [below=of S6] {\CIII{} \\ Reduced coords};
	\node[derivedspace] (P3) [below=of H6] {\PIII{} \\ Polar coords};

	% Arrows from F3 to row 2
	\draw[->, thick] (F3) -- (G6);
	\draw[->, thick] (F3) -- (S6);
	\draw[->, thick] (F3) -- (H6);

	% Arrows from row 2 to row 3
	\draw[->, thick] (G6) -- (DC7u);
	\draw[->, thick] (S6) -- (C3);
	\draw[->, thick] (H6) -- (P3);

\end{tikzpicture}
\label{fig_spaces}
\caption{Schematic relationships among crystallographic spaces. The base vector 
	space \FIII{} gives rise to three scalar spaces: metric tensor space \GVI{}, 
	Selling/Delone space \SVI{}, and conventional cell parameter space \HVI{}. 
	Each transforms into a reduced or classification space: $\mathbf{DC7u}$ from \GVI{}, 
	\CIII{} from \SVI{}, and \PIII{} from \HVI{}.}
\end{figure}

@@

@@

The first use applies the two-dimensional character of the coordinates of \PIII{}. They
allow simple-to-plot representations of collections of unit cells. Plotting
each of the coordinates of \PIII{} generates plots that are easily
interpreted as crystallographic axial length/opposite angle pairs. The
distance of a point has a length equal to the axial length, and the interaxisl
angle of the two other bases is illustrated on the plots. See Section  \ref{graphical}. Basically, together the three plots of the three coordinates are
simple projections of \HVI{}. See Section \ref{graphical}.

@@

While \PIII{} is derived from the same foundational cell parameters as $\mathbf{H^6}$, $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in the conventional cell representation. Compared with $\mathbf{C^3}$ \cite{andrews2023delone}, which derives coordinates from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic. It provides a useful intermediary: more interpretable than metric tensors or other spaces, but less abstract than \SVI{} or \GVI{}.

\section{Graphical display of projections}{
\label{graphical}
\PIII{}
allows simple-to-plot representations of collections of unit cells. 
See Figures \ref{PlotPolar_raw}, \ref{PlotPolar_Niggli}, and  \ref{PlotPolar_Delone} where 100 randomly chosen triclinic cells are plotted
(Figure \ref{PlotPolar_raw}) and the effects of unit cell reduction
are demonstrated (Figure \ref{PlotPolar_Niggli} and Figure \ref{PlotPolar_Delone} ).
}

\begin{figure}
    \begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_raw.png}
	\label{PlotPolar_raw}
	\caption{Scatter plots of 100 randomly selected
		 primitive triclinic unit cells represented in \PIII{}. Each of the 3 components of \PIII{}
		 is shown.}
	\end{center}
\end{figure}

\begin{figure}
    \begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_Niggli.png}
	\label{PlotPolar_Niggli}
	\caption{The same 100 random primitive triclinic unit cells shown in Figure \ref{PlotPolar_raw}, Niggli-reduced}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_Delone.png}
	\label{PlotPolar_Delone}
	\caption{Scatter plots of 100 randomly selected
		primitive triclinic unit cells represented in \PIII{} as shown in Figure \ref{PlotPolar_raw}, Selling (Delaunay)-reduced}
	\end{center}
\end{figure}

@@

In three-space, measuring the Euclidean distance between two points is just computing the square root
of the sum of the squares of the differences in the three coordinates. Similarly in \PIII{},
we compute distances from the sum of squared difference between the three
polar coordinates. But polar coordinates have a Euclidean measure from
their xy representation. So taking the six scalar values of two unit cells
presented in \PIII{}, we calculate a Euclidean distance by the ordinary
method.

%; CmdDists, calculate distances
%; Enter control variables and input vectors (type 'end' to finish):
%P 10 10 10  90 90 90
%p 10 10 10  90 90 91
%p 10 10 10.1 90 90 90
%end
%10.000    10.000    10.000    90.000    90.000    90.000
%14.142    30.113     8.479   147.466    56.504   144.893
%
%10.000    10.000    10.000    90.000    90.000    90.000
%14.142    29.916     8.564   147.586    56.504   144.621
%
%<pre style="font-size: 105%; line-height: 60%; display:inline;" >;
%; ordinals   P3-Euclid    P3Dist       NCDist    || ordinals     P3-Euclid    P3Dist       NCDist
%; 1-2        0.175      19191.000      3.490         1-2         0.175      19191.000      3.490
%; 2-3        0.202      19191.000      4.028         1-3         0.100      19191.000      2.010
%
%
%;.......... input.cells ........................................ ordinals .....
%P   10.000    10.000    10.000    90.000    90.000    90.000     1
%P   10.000    10.000    10.000    90.000    90.000    91.000     2
%P   10.000    10.000    10.100    90.000    90.000    90.000     3
%;</pre>

%; CmdDists, calculate distances
%; Enter control variables and input vectors (type 'end' to finish):
%p 10 10 10   90 90 90
%p 10 10 10   90 90 91
%p 10 10 11   90 90 90
%end
%<pre style="font-size: 105%; line-height: 60%; display:inline;" >;
%; ordinals   P3-Euclid    P3Dist       NCDist    || ordinals     P3-Euclid    P3Dist       NCDist
%; 1-2        3.490        0.203        3.490         1-2         0.175       26.959        3.490
%; 2-3        21.288       0.872       21.288         1-3         1.000       26.896       21.000
%
%
%;.......... input.cells ........................................ ordinals .....
%P   10.000    10.000    10.000    90.000    90.000    90.000     1
%P   10.000    10.000    10.000    90.000    90.000    91.000     2
%P   10.000    10.000    11.000    90.000    90.000    90.000     3
%;</pre>

Let us compare two cells derived by modifying a primitive unit cell to form two slightly modified cells, one by changing $\gamma$, and one by changing $c$.
\begin{enumerate}
	\item P 10 10 10  90 90 90
	\item P 10 10 10  90 90 91
	\item P 10 10 10.1  90 90 90
\end{enumerate}
The Euclidean distances between these cells are:
\begin{itemize}
	\item 1-2: 0.175
	\item 1-3: 0.100
	\item 2-3: 0.202
\end{itemize}

In this example, a 1.0° change in $\gamma$ yields a Euclidean distance comparable to a 0.1 Å change in the c-axis length.
%
%\section{Scalar Reduction and Canonical Presentation in Projected Vector Triplets}
%
%\subsection{1. Standard Presentation of Projected Vectors}
%	
%	Let a set of three projected vectors be given:
%	
%	
%	\[
%	(p_1, p_2, p_3), \quad p_i \in \mathbb{R}^2
%	\]
%	
%	
%	Each \( p_i \) is constructed from a lattice vector via magnitude and opposing angle:
%	
%	
%	\[
%	p_i = |\vec{v}_i| (\cos \theta_i, \sin \theta_i)
%	\]
%	
%	
%	
%	The \textbf{Standard Presentation} algorithm assigns a canonical ordering and orientation to the triplet:
%	
%	\begin{enumerate}
%		\item \textbf{Magnitude Sort:} Reorder vectors so that
%		
%		
%		\[
%		|p_1| \leq |p_2| \leq |p_3|
%		\]
%		
%		
%		using Euclidean norms \( |p_i| = \sqrt{x_i^2 + y_i^2} \). Ties are preserved up to a numerical tolerance \( \delta \).
%		
%\item \textbf{Directional Coherence:} If the \( x \)-components of the triplet are not all of consistent sign, flip the necessary vectors by negating only the \( x_i \) component:
%
%
%\[
%p_i \to (-x_i,\ y_i)
%\]
%
%
%This maps quadrant I vectors into quadrant II, preserving their placement in the upper half-plane and maintaining angular conventions typical in crystallographic analysis.
%
%
%
%\[
%p_i \to (-x_i,\ y_i)
%\]
%
%
%to align the triplet into a directionally consistent state. If all \( x_i < 0 \), the configuration is already coherent and requires no change.
%
%		
%		
%		\[
%		p_i \to (-x_i, y_i)
%		\]
%		
%		
%		for any two vectors, preserving chirality.
%		
%		\item \textbf{Return Ordered Triplet:} Output the permuted and flipped triplet with consistent magnitude order and cosine directionality.
%	\end{enumerate}
%	
%	\vspace{1em}
%	
%	\subsection{2. Scalar Projection Interaction and Cost Function}
%	
%	For any pair of projected vectors \( p_i, p_j \in \mathbb{R}^2 \), define the scalar projection interaction (or directional affinity):
%	
%	
%	\[
%	\mu_{ij} = p_i \cdot p_j = x_i x_j + y_i y_j
%	\]
%	
%	
%	
%	The \textbf{scalar coupling cost} is then defined over the triplet as:
%	
%	
%	\[
%	C' = \frac{|\mu_{12}| + |\mu_{13}| + |\mu_{23}|}{p_1 \cdot p_1 + p_2 \cdot p_2 + p_3 \cdot p_3}
%	\]
%	
%	
%	
%	This cost measures the normalized sum of pairwise directional alignment. Lower values of \( C' \) indicate that the projected vectors are less directionally coupled — i.e., more orthogonal or dispersed.
%	
%	\vspace{1em}
%	
%	\subsection{3. Scalar Reduction Algorithm}
%	
%	The goal of scalar reduction is to transform the triplet \( (p_1, p_2, p_3) \) into an algebraically simpler form — one with lower scalar coupling cost \( C' \) — while preserving lattice identity (e.g., volume and reconstructability).
%	
%	The reduction proceeds iteratively:
%	
%	\begin{enumerate}
%		\item \textbf{Initialization:}
%		Apply standard presentation. Compute initial cost \( C'_0 \) and physical volume \( V_0 > 0 \).
%		
%		\item \textbf{Pairwise Projection Subtractions:}
%		For each unordered pair \( (p_i, p_j),\ i \neq j \), compute the scalar coefficient:
%		
%		
%		\[
%		\lambda = \frac{p_i \cdot p_j}{p_j \cdot p_j}
%		\]
%		
%		
%		Then define the candidate vector:
%		
%		
%		\[
%		p_i' = p_i - \lambda p_j
%		\]
%		
%		
%		
%		\item \textbf{Evaluate Candidate:}
%		Apply standard presentation to updated triplet. Recompute cost \( C'_{\text{new}} \) and volume \( V_{\text{new}} \).
%		
%		\item \textbf{Acceptance Criteria:}
%		Accept the transformation if:
%		
%		
%		\[
%		C'_{\text{new}} + \delta < C' \quad \text{and} \quad V_{\text{new}} > 0
%		\]
%		
%		
%		
%		\item \textbf{Repeat:}
%		Continue until no further accepted transformations lower the cost.
%		
%		\item \textbf{Output:}
%		Return the reduced triplet \( (p_1, p_2, p_3)_{\text{red}} \) with minimal scalar cost.
%	\end{enumerate}
%	
%	\vspace{1em}
%	
%	\subsection{4. Interpretation of Cost and Reduction}
%	
%	The term "reduction" refers to minimizing the scalar coupling among vectors — reducing directional alignment that obscures presentation clarity. The cost \( C' \) is a rigorously defined metric for this alignment. Transformations reduce \( C' \) while preserving:
%	\begin{itemize}
%		\item The vector magnitudes and their reconstructable relationships
%		\item The unit cell's volume and crystallographic identity
%		\item The triplet’s algebraic validity in scalar projection space
%	\end{itemize}
%	
%	Reduction leads to a cleaner, more canonical representation of the original structure — just as Niggli reduction provides a minimal cell in metric tensor space, scalar reduction provides a minimal configuration in projected vector space.
%	
%

@@

	The space \PIII{} is introduced. Less abstract than space based
	on dot products and metric tensors, it provides a simple to understand
	method for comparing unit cells and a convenient method of displaying 
	unit cells.

@@

	%\section{blah blah blah -- Supplementary Material}
	\ack{{\bf Acknowledgements}}

	Careful copy-editing and corrections by Frances C. Bernstein are 
	gratefully acknowledged.
	%	Our thanks to Jean Jakoncic and Alexei Soares for 
	%	helpful conversations and access to data and facilities at 
	%	Brookhaven National Laboratory.
	%	
	\ack{{\bf Funding information}}

	Funding for this research was provided in part by:  
	US Department of Energy Offices of Biological and 
	Environmental Research and of Basic Energy Sciences 
	(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
	U.S. National Institutes of Health (grant No. P41RR012408; 
	grant No. P41GM103473; grant No. P41GM111244; 
	grant No. R01GM117126,
	grant No. 1R21GM129570); Dectris, Ltd.

@@

	\PIII{} is a linearized description of unit cell parameters,
	derived from usual lengths and angles of the unit cell. It 
	provides a useful metric for comparing unit cells and a
	convenient base for displaying and comparing unit cells.

>>>>

	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
	\author[a]{Herbert J.}{Bernstein}

\aff[a]{Ronin Institute for Independent Scholarship 2.0, USA}

@@

	\begin{abstract}
		The space \PIII{} is introduced, derived from unit cell axial lengths and interaxial angles. \PIII{} enables linearization of unit cell parameters using three polar coordinate bases.

@@

[\underline{	Note: Boris Delaunay in his later publications chose to render
	his surname as ``Delone'' (which is still pronounced the same as Delaunay).}]

	For the above reasons, other spaces are often used to describe lattices, often for specialized purposes; see Table  \ref{tbl_spaces}. Figure \ref{fig_spaces} shows relationships among
	several of these spaces. Although some of these spaces, such as \HVI{} and	\FIII{} are in almost daily use, they are not often referred to 
	as ``spaces''. In the case of \FIII{}, symmetry elements are linear
	operations in this space, and fractional coordinates of atomic positions
	are vectors expressed in this space.

@@

	} &
	Composite geometric parameter space $\mathbb{R}_{>0}^3 \times (0^\circ, 180^\circ)^3$. Lengths are unconstrained; angles are geometrically restricted. The angles must obey the spherical triangle inequalities (summing to less than 360 degrees, and the sum of any 2 must not exceed the third). \HVI{} is useful for conventional lattice descriptions, but limited in symmetry representation and metric-based classification. \\
	\midrule
	\FIII{} &
	$\vec{a},\, \vec{b},\, \vec{c} \in \mathbb{R}^3$ &
	Base vector space in \RIII. Coordinates are the three-space base vectors of a unit cell. Lattices are defined using the
	components of \FIII{}, and translational symmetry is described 
	in \FIII{}. \\	
	\midrule

@@

	\PIII{} &
	\parbox[t]{5.5cm}{
		$(|\vec{a}|,\alpha),\ (|\vec{b}|,\beta),\ (|\vec{c}|,\gamma)$\\
		$\Rightarrow$\\
		\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha)$\\
		\hspace*{0.5cm}$(|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta)$\\
		\hspace*{0.5cm}$(|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$\\
		also known as $(p_1, p_2, p_3)$
	} &
	Polar coordinate space defined in terms of unit cell axial
	lengths and their opposing angles (opposing, the angle between the other two unit cell axes). It is a compact and smooth alternative to $\HVI$. Useful for comparing unit cells. (This work) \\
		\midrule

	\SVI{} &
	\parbox[t]{5.5cm}{
		$s_1 = \vec{b} \cdot \vec{c},\, s_2 = \vec{a} \cdot \vec{c},\, s_3 = \vec{a} \cdot \vec{b}$\\
		$s_4 = \vec{a} \cdot \vec{d},\, s_5 = \vec{b} \cdot \vec{d},\, s_6 = \vec{c} \cdot \vec{d}$\\
		\quad where $\vec{d} = -\vec{a} - \vec{b} - \vec{c}$
	} &
	Used for Selling/Delone reduction. The scalar components are dot products between lattice vectors. \SVI{} supports Bravais lattice classification. 24 Bravais types were defined using the scalar components of \SVI{} by Delone (1975), later revised using \SVI{} \cite{andrews2023measuring}. \\
		\midrule

@@

\CIII{} &
	\parbox[t]{5.5cm}{
		From \SVI{} scalars: $s_1,\dots,s_6$\\
		$c_1 = (s_1, s_4),\ c_2 = (s_2, s_5),\ c_3 = (s_3, s_6)$\\
		where the parentheses denote a complex number
	} &
	\CIII{} associates scalar pairs identified by Delone as ``opposites.'' Symmetry and Selling/Delone reduction simplify in this space \cite{andrews2023complex}. \\
		\midrule

@@

\begin{figure}
\includegraphics[]{P3_fig1.png}
\label{fig_spaces}
\caption{Schematic relationships among crystallographic spaces. The base vector 
	space \FIII{} gives rise to three scalar spaces: metric tensor space \GVI{}, 
	Selling/Delone space \SVI{}, and conventional cell parameter space \HVI{}. 
	Each transforms into a reduced or classification space: $\mathbf{DC7u}$ from \GVI{}, 
	\CIII{} from \SVI{}, and \PIII{} from \HVI{}.}
\end{figure}

@@

As described in Table \ref{tbl_spaces}, the Selling scalars are a vector:
[$s_1, s_2, s_3, s_4, s_5, s_6$]. \citeasnoun{Delaunay1932} represents
their relationships as the edge components of a tetrahedron as 
shown in Figure \ref{tetrahedron}.

\begin{figure}
	\begin{center}
		\includegraphics[width=0.3\columnwidth]{tetrahedron}
		\label{tetrahedron}
		\caption{The relationships of the Selling scalars as
		described by  \citeasnoun{Delaunay1932}, but labeled using the vector components of \SVI{} \cite{andrews2023measuring}. Visualized as a
		3 dimensional tetrahedron, scalar $s_1$ is
		opposite scalar $s_4$, etc. The number of each scalar is the 
		position in the \SVI{} vector. See Table \ref{tbl_spaces}
		for the definitions of the scalars in space \SVI{}. }

	\end{center}
\end{figure}

@@

The first use applies the two-dimensional character of the coordinates of \PIII{}. They
allow simple-to-plot representations of collections of unit cells. Plotting
each of the coordinates of \PIII{} generates plots that are easily
interpreted as crystallographic axial length/opposite angle pairs. The
distance of a point from the origin has a length equal to the axial length, and the interaxial
angle between the 2 other bases (plotted as the angle from the x-axis) is illustrated on the plots. Basically, together the three plots of the three coordinates are
simple projections of \HVI{}. See Section \ref{graphical}.

@@

While \PIII{} is derived from the same foundational cell parameters as $\mathbf{H^6}$, $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in the conventional cell representation. Compared with $\mathbf{C^3}$ \cite{andrews2023complex}, which derives coordinates from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic. It provides a useful intermediary: more interpretable than metric tensors or other spaces, but less abstract than \SVI{} or \GVI{}.

\section{Graphical display of projections}{
\label{graphical}
\PIII{}
allows simple-to-plot representations of collections of unit cells. 
See Figures \ref{PlotPolar_raw}, \ref{PlotPolar_Niggli}, and  \ref{PlotPolar_Delone}, where 100 randomly chosen triclinic cells are plotted
(Figure \ref{PlotPolar_raw}), and the effects of unit cell reduction
are demonstrated (Figure \ref{PlotPolar_Niggli} and Figure \ref{PlotPolar_Delone} ).
}

\begin{figure}
    \begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_raw.png}
	\label{PlotPolar_raw}
	\caption{Scatter plots of 100 randomly selected
		 primitive triclinic unit cells represented in \PIII{}. Each of the 3 components of \PIII{}
		 is shown. (Colors reflect the order of input of the unit cells.)}
	\end{center}
\end{figure}

\begin{figure}
    \begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_Niggli.png}
	\label{PlotPolar_Niggli}
	\caption{The same 100 randomly selected primitive triclinic unit cells shown in Figure \ref{PlotPolar_raw}, Niggli-reduced}.
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_Delone.png}
	\label{PlotPolar_Delone}
	\caption{Scatter plots of 100 randomly selected
		primitive triclinic unit cells represented in \PIII{} as shown in Figure \ref{PlotPolar_raw}, Selling /Delone-reduced}.
	\end{center}
\end{figure}

@@

In 3-space, measuring the Euclidean distance between two points is just computing the square root
of the sum of the squares of the differences in the three coordinates. Similarly in \PIII{},
we compute distances from the square root of the sum of squared differences between the three
polar coordinates. Polar coordinates have a Euclidean measure from
their xy representation. So, taking the 6 scalar values of 2 unit cells
presented in \PIII{}, we calculate a Euclidean distance by the ordinary
method.

Let us compare two cells derived by modifying a primitive unit cell to form two slightly modified cells, one by changing $\gamma$, and one by changing $\vec{c}$.
\begin{enumerate}
	\item P 10 10 10  90 90 90
	\item P 10 10 10  90 90 91
	\item P 10 10 10.1  90 90 90
\end{enumerate}
The Euclidean distances between these cells are:
\begin{itemize}
	\item 1--2: 0.175
	\item 1--3: 0.100
	\item 2--3: 0.202
\end{itemize}

In this example, a 1.0$^\circ$ change in $\gamma$ yields a Euclidean distance comparable to a 0.1~\AA{} change in the $c$-axis length.

@@

	The space \PIII{} is introduced. Less abstract than spaces based
	on dot products and metric tensors, it provides a simple way to understand the
	method for comparing unit cells and a convenient method of displaying 
	unit cells.

@@

	\ack{{\bf Acknowledgements}}

Careful copy-editing and corrections by Frances C. Bernstein are 
gratefully acknowledged.

	\ack{{\bf Funding information}}

	Funding for this research was provided in part by:  
	U.S. Department of Energy Offices of Biological and 
	Environmental Research and of Basic Energy Sciences 
	(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
	U.S. National Institutes of Health (grant No. P41RR012408; 
	grant No. P41GM103473; grant No. P41GM111244; 
	grant No. R01GM117126,
	grant No. 1R21GM129570); Dectris, Ltd.

@@

	\PIII{} is a linearized description of unit cell parameters,
	created from the usual lengths and angles of the unit cell. It 
	provides a useful metric for comparing unit cells and a
	convenient base for displaying and comparing unit cells.
